Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2024-12-12 manuscript/paper.tex
Wrap lines

\abstract{
%\textbf{Motivation:}
Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic ancestry for a set of recombining sequences. The ability to capture the evolutionary history of samples makes ARGs valuable in a wide range of applications in population and statistical genetics. ARG-based approaches are increasingly becoming a part of genetic data analysis pipelines due to breakthroughs enabling ARG inference at biobank-scale. However, there is a lack of visualisation tools, which are crucial for validating inferences and generating hypotheses. 
%\textbf{Results:} 
We present \texttt{tsbrowse}, an open-source Python web-app for the interactive visualisation of the fundamental building-blocks of ARGs, i.e.,  nodes, edges and mutations. We demonstrate the application of \texttt{tsbrowse} to various data sources and scenarios, and highlight its key features of browsability along the genome, user interactivity, and scalability to very large sample sizes.\\
\textbf{Availability:}\\ Python package: \texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version: \texttt{\url{https://github.com/tskit.dev/tsbrowse}}, \\
Documentation: \texttt{\url{https://tskit.dev/tsbrowse/docs/}}\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Available at \textit{\textcolor{red}{\href{https://www.google.com}{Bioinformatics} online}.}}
}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/MainFig1.png}
    \caption{\textbf{Overview of tsbrowse architecture. } 
    In tskit, ARGs are defined by a collection of tables which broadly map to ARG properties. Exemplar table names are denoted in dark blue and columns that describe the property in light blue. In the pre-process step, tables are augmented with additional information computed for each property (yellow). The output from this pre-processing step is stored as a \texttt{.tsbrowse} file. Next the serve step renders the visualisation in a web browser by leveraging tools in the Holoviz ecosystem. Optional annotations are provided as an input file, allowing users to overlay contextual information about genes or other sequence features. 
    }
    \label{fig:Figure_1}
\end{figure}

@@

\subsection{User Interface}
\label{subsec:User_Interface}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/MainFig2.png}
    \caption{\textbf{A view of ARG edges in \texttt{tsbrowse}.}
    A screenshot of the Edges tab for a 5 Mb sequence simulated with selection coefficient, \textit{s = 0.25}, and beneficial allele situated at \textit{x = 2.5 Mb}. The plot on the left represents each edge as a horizontal line along the sequence (X-axis). The time of parent node is shown on the Y-axis. The persistence of edges across various dimensions is shown through histograms on the right: top, distribution of genomic spans; middle, distribution of time spans; bottom, edge area, which is the product of genomic and time spans.} 
    \label{fig:Figure_2}
\end{figure}

@@

The user interface is presented as a \texttt{Panel} dashboard, with pages to describe mutations, nodes, edges and trees in the input ARG. Figure 2 demonstrates the user interface using the Edges page as an example. The plot on the left is an overview of all 33,929 edges (parent-child relationships between node-pairs; see \hyperref[subsec:Data_Model]{Data Model}) in a simulated ARG \hyperref[subsec:sweep_simulation]{described in Supplementary Methods}; each edge is depicted as a horizontal line connecting the genomic coordinates of the parent and child nodes on the x-axis. The y-axis denotes the age of either node. A useful feature of the app is the ability to interact with the visualisations. The tools on the right of the plot allow the user to examine edges within a sequence/time slice, and query individual edges. The histograms to the right provide summaries of edge spans over sequence and time.
%The ability to explore ARG properties across the sequence and interact with the data makes the analysis of local genealogical patterns more intuitive.

@@

\section*{Acknowledgments}
\section*{Author contributions}
\section*{Supplementary Data}
Supplementary data are available at \textcolor{red}{Bioinformatics online}.
\section*{Conflict of interest}
No competing interest is declared.
\section*{Funding}
This work was supported by funding from the Biotechnology and Biological Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1], \textcolor{red}{add Novo funding and funding details for Ben, Duncan, Jerome}.
\bibliographystyle{natbib}
\bibliography{manuscript/reference}

@@

\clearpage
\section{Supplementary Figures}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig1.png}
    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate (A), ARG-Needle (B), Relate (C) and SINGER (4) inferences of the truth dataset simulated under a selective sweep model (in Figure 2 of the main text). For SINGER, an examplar ARG from the set of output posterior ARG samples is shown. The X coordinate represents genomic position, each horizontal segment on the plot shows the genomic coordinates that the edge spans, and Y coordinate shows time of either the parent or child node in the edge.}
    \label{fig:Supplementary_Figure_1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig2.png}
    \caption{\textbf{A view of mutations in the 1000G WGS dataset.}
    A screenshot of \texttt{tsbrowse}'s Mutations view for a X Mb region on chromosome 17 inferred from from 3,202 participants from the 1000 Genomes Whole Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}. The genomic position of the mutations is shown on the X axis, and the age of mutation is represented on the Y axis. Panel B represents an inference after filtering out low site-density regions in the samples.}
    \label{fig:Supplementary_Figure_2}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig3.png}
    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.}
    A screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm of chromosome 20 from the 1000 Genomes whole-genome sequencing dataset. At the top is a plot of node spans over time. The length of sequence that the nodes span is shown on the X axis, and the time of nodes is shown on the Y axis. The histogram at the bottom shows the distribution of node spans. }
    \label{fig:Supplementary_Figure_3}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig4.png}
    \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV2 ARGs.}
    A screenshot of \texttt{tsbrowse}'s depiction of 988,303 mutations in an 
    ARG inferred by \texttt{sc2ts}; see  text for details. Also shown are the 
    gene annotations along the X-axis.}
    \label{fig:Supplementary_Figure_4}
\end{figure}
\end{document}

>>>>

\abstract{
%\textbf{Motivation:}
Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic
ancestry for a set of recombining sequences. The ability to capture the
evolutionary history of samples makes ARGs valuable in a wide range of
applications in population and statistical genetics. ARG-based approaches are
increasingly becoming a part of genetic data analysis pipelines due to
breakthroughs enabling ARG inference at biobank-scale. However, there is a lack
of visualisation tools, which are crucial for validating inferences and
generating hypotheses. 
%\textbf{Results:} 
We present \texttt{tsbrowse}, an open-source Python web-app for the interactive
visualisation of the fundamental building-blocks of ARGs, i.e.,  nodes, edges
and mutations. We demonstrate the application of \texttt{tsbrowse} to various
data sources and scenarios, and highlight its key features of browsability
along the genome, user interactivity, and scalability to very large sample
sizes.\\ \textbf{Availability:}\\ Python package:
\texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version:
\texttt{\url{https://github.com/tskit.dev/tsbrowse}}, \\ Documentation:
\texttt{\url{https://tskit.dev/tsbrowse/docs/}}\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Available at
%\textit{\textcolor{red}{\href{https://www.google.com}{Bioinformatics}
%online}.}}
}

@@

\begin{figure} \centering
\includegraphics[width=0.95\linewidth]{figures/MainFig1.png}
\caption{\textbf{Overview of tsbrowse architecture. } In tskit, ARGs are
defined by a collection of tables which broadly map to ARG properties. Exemplar
table names are denoted in dark blue and columns that describe the property in
light blue. In the pre-process step, tables are augmented with additional
information computed for each property (yellow). The output from this
pre-processing step is stored as a \texttt{.tsbrowse} file. Next the serve step
renders the visualisation in a web browser by leveraging tools in the Holoviz
ecosystem. Optional annotations are provided as an input file, allowing users
to overlay contextual information about genes or other sequence features. }
\label{fig:Figure_1} \end{figure}

@@

\subsection{User Interface}
\label{subsec:User_Interface}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
    \caption{\textbf{A view of ARG edges in \texttt{tsbrowse}.}
    A screenshot of the Edges tab for a 5 Mb sequence simulated with selection coefficient, \textit{s = 0.25}, and beneficial allele situated at \textit{x = 2.5 Mb}. The plot on the left represents each edge as a horizontal line along the sequence (X-axis). The time of parent node is shown on the Y-axis. The persistence of edges across various dimensions is shown through histograms on the right: top, distribution of genomic spans; middle, distribution of time spans; bottom, edge area, which is the product of genomic and time spans.}

    \label{fig:Figure_2}
\end{figure}

@@

The user interface is presented as a \texttt{Panel} dashboard, with pages to
describe mutations, nodes, edges and trees in the input ARG. Figure 2
demonstrates the user interface using the Edges page as an example. The plot on
the left is an overview of all 33,929 edges (parent-child relationships between
node-pairs; see \hyperref[subsec:Data_Model]{Data Model}) in a simulated ARG
\hyperref[subsec:sweep_simulation]{described in Supplementary Methods}; each
edge is depicted as a horizontal line connecting the genomic coordinates of the
parent and child nodes on the x-axis. The y-axis denotes the age of either
node. A useful feature of the app is the ability to interact with the
visualisations. The tools on the right of the plot allow the user to examine
edges within a sequence/time slice, and query individual edges. The histograms
to the right provide summaries of edge spans over sequence and time.

%The ability to explore ARG properties across the sequence and interact with the data makes the analysis of local genealogical patterns more intuitive.

@@

\section*{Acknowledgments}
\section*{Author contributions}
\section*{Supplementary Data}
Supplementary data are available at \textcolor{red}{Bioinformatics online}.
\section*{Conflict of interest}
No competing interest is declared.
\section*{Funding}
This work was supported by funding from the Biotechnology and Biological
Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1],
\textcolor{red}{add Novo funding and funding details for Ben, Duncan, Jerome}.
\bibliographystyle{natbib}
\bibliography{reference}

@@

\clearpage
\section{Supplementary Figures}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig1.png}
    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate (A),
ARG-Needle (B), Relate (C) and SINGER (4) inferences of the truth dataset
simulated under a selective sweep model (in Figure 2 of the main text). For
SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
The X coordinate represents genomic position, each horizontal segment on the
plot shows the genomic coordinates that the edge spans, and Y coordinate shows
time of either the parent or child node in the edge.}
    \label{fig:Supplementary_Figure_1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
    \caption{\textbf{A view of mutations in the 1000G WGS dataset.} A
screenshot of \texttt{tsbrowse}'s Mutations view for a X Mb region on
chromosome 17 inferred from from 3,202 participants from the 1000 Genomes Whole
Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}. The genomic
position of the mutations is shown on the X axis, and the age of mutation is
represented on the Y axis. Panel B represents an inference after filtering out
low site-density regions in the samples.}
    \label{fig:Supplementary_Figure_2}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig3.png}
    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.} A
screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm
of chromosome 20 from the 1000 Genomes whole-genome sequencing dataset. At the
top is a plot of node spans over time. The length of sequence that the nodes
span is shown on the X axis, and the time of nodes is shown on the Y axis. The
histogram at the bottom shows the distribution of node spans. }
    \label{fig:Supplementary_Figure_3}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig4.png}
    \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV2 ARGs.}
    A screenshot of \texttt{tsbrowse}'s depiction of 988,303 mutations in an 
    ARG inferred by \texttt{sc2ts}; see  text for details. Also shown are the 
    gene annotations along the X-axis.}
    \label{fig:Supplementary_Figure_4}
\end{figure}
\end{document}
